### 2020-01-30 nlp_models.tex
mwm initial pass

@@

In the past two years, nearly 100 open source, pre-trained DNNs for Natural Language Processing (NLP) have emerged,
based on the revolutionary Transformer architecture, including variants of BERT, Transformer-XML, GPT, etc.
The Transformer architectures consist of blocks of Attention layers, containing of 2 large, Feed Forward (Linear)
weight matrics \cite{Attn2017}. In contrast to the smaller pre-Activation maps arising in Cond2D layers,
the Attention matrices are significantly larger, and frequently under-correlated, with generally larger
Heavy Tailed Power Law exponents $\alpha$ (when fitting the ESD).  Here, we breifly look at a few
popular pretrained NLP DNNs to demonstrate that the Theory of Heavy Tailed Self-Regularization also
applies to NLP models and not just CV models, and to highlight how to use the theory to identify
poorly trained models where the empirical norm metrics will perform poorly.

\paragraph{OpenAI GPT Models}
We use the \emph{WeightWatcher} tool to analyze the OpenAI GPT and GPT2 models, which gives us
the opportunity to analyze the effect of both training the same model with different size data sets,
and increasing sizes of both the data set and architectures.
These models have generated significant media attention because of their remarkable ability to
 generate fake text that appears to be real and the potential misuse of this.
For this reason, the original GPT model was trained on on a deficient data set, rendering
the model interesting but not fully functional.  Later, OpenAI released a much improved model--
GPT2 (small)--which has the same architecture and number of layers as GPT, but has
been trained on a larger and better data set (and with other changes), making it
remarkably good at generating (near) human-quality fake text.  
By comparing the poorly trained GPT to the well trained GPT2, we
indentify empirical indidcators for when  model has in fact
been poorly trained and may perform poorly when deployed.

@@

\paragraph{Empirical Quality Metrics for GPT and GPT2}
We first compare the distribution of  Heavy Tailed Power Law exponents $\alpha$ in GPT and GPT2.
They are very differenmt, with GPT2 having both a notably smaller mean $\alpha$, and far fewer, unusually large outliers.
Figure \ref{fig:gpt-alphs-hist} shows the empirical density (histogram) of $\alpha$
for all layers in GPT (blue) and GPT2 (red).  \charles{discuss more}

@@

\paragraph{The Correlation Flow in GPT and GPT2} also differs significantly between GPT and GPT2.
Figure \ref{fig:gpt-alpha-layer} plots $\alpha$ vs the depth (i.e. a layer id) for each model.

@@

We do notice, however, that while the peaks of the $\alpha$ is getting smaller, towards $2.0$,
the tails of the distribution shifts right, with larger GPT2 models having more
usually large $\alpha$.  We suspect this indicates that these larger GPT2 models are 
overparameterized and/or not yet fully optimized and require datasets even larger
than the recent XL $1.5B$ release\cite{gpt2-xl}.

>>>>

In this section, we examine the quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.

apply more generally, where they differ, e.g., how the layer Spectral Norm behaves unexpectedly, while the $\alpha$-based metrics can give insight into how well trained a model is.

@@

Within the past two years, nearly 100 open source, pre-trained DNNs for NLP have emerged, based on the revolutionary Transformer architecture.
This includes variants of BERT, Transformer-XML, GPT, etc.
The Transformer architectures consist of blocks of Attention layers, containing two large, Feed Forward (Linear) weight matrics~\cite{Attn2017}. 
In contrast to the smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger, 
and frequently under-correlated, with generally larger Heavy Tailed Power Law exponents $\alpha$ (when fitting the ESD).  
Here, we breifly look at a few popular pretrained NLP DNNs to demonstrate that the Theory of Heavy Tailed Self-Regularization also
applies to NLP models and not just CV models, and to highlight how to use the theory to identify
poorly trained models where the empirical norm metrics will perform poorly.

\paragraph{OpenAI GPT Models.}

We use the \emph{WeightWatcher} tool to analyze the OpenAI GPT and GPT2 models, which gives us
the opportunity to analyze the effect of both training the same model with different size data sets,
and increasing sizes of both the data set and architectures.
These models have generated significant media attention because of their remarkable ability to
 generate fake text that appears to be real and the potential misuse of this.
For this reason, the original GPT model was trained on on a deficient data set, rendering
the model interesting but not fully functional.  Later, OpenAI released a much improved model--
GPT2 (small)--which has the same architecture and number of layers as GPT, but has
been trained on a larger and better data set (and with other changes), making it
remarkably good at generating (near) human-quality fake text.  
By comparing the poorly trained GPT to the well trained GPT2, we
indentify empirical indidcators for when  model has in fact
been poorly trained and may perform poorly when deployed.

@@

\paragraph{Empirical Quality Metrics for GPT and GPT2.}

We first compare the distribution of  Heavy Tailed Power Law exponents $\alpha$ in GPT and GPT2.
They are very differenmt, with GPT2 having both a notably smaller mean $\alpha$, and far fewer, unusually large outliers.
Figure \ref{fig:gpt-alphs-hist} shows the empirical density (histogram) of $\alpha$
for all layers in GPT (blue) and GPT2 (red).  \charles{discuss more}

@@

\paragraph{Correlation Flow in GPT and GPT2.}

also differs significantly between GPT and GPT2.
Figure \ref{fig:gpt-alpha-layer} plots $\alpha$ vs the depth (i.e. a layer id) for each model.

@@

We do notice, however, that while the peaks of the $\alpha$ is getting smaller, towards $2.0$,
the tails of the distribution shifts right, with larger GPT2 models having more
usually large $\alpha$.  We suspect this indicates that these larger GPT2 models are 
overparameterized and/or not yet fully optimized and require datasets even larger
than the recent XL $1.5B$ release~\cite{gpt2-xl}.
